---
title: "Computer Lab 5 Computational Statistics"
author: "Phillip Hölscher & Zijie Feng"
date: "23-2-2019"
output: 
  pdf_document:
    toc: true 
---

```{r setup, include=FALSE}
Sys.setlocale(locale = "English")
knitr::opts_chunk$set(echo = TRUE, out.height = "270px")
# libraries used in this lab
library(ggplot2)
#install.packages("boot") # to create booststarp
library(boot)
```

\newpage

# Question 1: Hypothesis testing

In 1970, the US Congress instituted a random selection process for the military draft. All 366 possible birth dates were placed in plastic capsules in a rotating drum and were selected one by one. The first date drawn from the drum received draft number one, the second date drawn received draft number two, etc. Then, eligible men were drafted in the order given by the draft number of their birth date. In a truly random lottery there should be no relationship between the date and the draft number. Your task is to investigate whether or not the draft numbers were randomly selected. The draft numbers (`Y=Draft_No`) sorted by day of year (`X=Day_of_year`) are given in the file *lottery.xls*.

## 1. Make a scatterplot 
of Y versus X and conclude whether the lottery looks random.

Scatterplot of Y versus X
```{r, echo=FALSE}
# scatterplot 
rm(list=ls())
data1 = read.csv2("lottery.csv")

plot_11 = ggplot(data = data1,aes(x = Day_of_year, y = Draft_No)) + 
  geom_point(aes(color = "red")) 
plot_11
```

Since the dots are spaced over the whole area, it seems like the military draft is truly random.

## 2. Compute an estimate 

$\hat{Y}$ of the expected response as a function of $X$ by using a loess smoother (use `loess()`), put the curve $\hat{Y}$ versus $X$ in the previous graph and state again whether the lottery looks random.

Scatterplot of $\hat{Y}$ versus X
```{r, echo=FALSE}
# loess(formula = Draft_No~Day_of_year, data = data1)

plot_11 + 
  # geom_smooth()` using method = 'loess' and formula 'y ~ x'
  geom_smooth(method ="loess", formula = y~x, aes(color="blue")) +
  scale_color_discrete(labels=c("expected response",
                                "scatter points")
                       )
```

It does look like the loess smoother decreases slowly after X (= Day of year) passes number 100. This could mean the draft is not truly random. 

## 3. Check whether the lottery is random
To check whether the lottery is random, it is reasonable to use test statistics

$$ T = \frac{\hat{Y}(X_b)-\hat{Y}(X_a)}{X_b - X_a}, ~~\rm where~~  X_b = \rm argmax_X Y(X), ~\  X_a = \rm argmin_X Y(X)  $$

If this value is significantly greater than zero, then there should be a trend in the data and the lottery is not random. Estimate the distribution of $T$ by using a non–parametric bootstrap with $B = 2000$ and comment whether the lottery is random or not. What is the p–value of the test?

```{r}
# use non-parametric bootstrap 
# need data = data1, statistic = loess & test stat, R = 2000
stats = function(data,vn){
  datatemp<-data[vn,]
  # Y=Draft_No & X=Day_of_year
  # create the loess regression based on the sample data
  reg = loess(Draft_No ~ Day_of_year, data = datatemp)
  # the X which owns the max/min Y in original data
  X_b = data$Day_of_year[which.max(data$Draft_No)]   
  X_a = data$Day_of_year[which.min(data$Draft_No)]
  # Y_hat(X)
  fit_X_b = predict(reg,newdata=X_b)
  fit_X_a = predict(reg,newdata=X_a)
  # the given test statistic
  T_stat = (fit_X_b - fit_X_a) / (X_b-X_a)
  return(T_stat)
}

#non-parametric bootstrap
set.seed(123456)
# dt=data1[order(data1$Draft_No),]
myboot = boot(data = data1, 
              statistic = stats,
              R = 2000)

# summary
myboot
# plot distribution
# plot(myboot, index = 1)
```
```{r echo=FALSE}
d <- data.frame(t=myboot$t)
mean_t = mean(myboot$t,na.rm = TRUE)
ggplot(data = d, aes(x = t)) + 
  ggtitle("Histogram of t") + 
  geom_histogram(aes(y=..density..),
                 colour="black",
                 fill="white",
                 bins=30) + 
  geom_vline(aes(xintercept = mean_t, color = "red"),size=1.5)
```

The test statistic*T* is around -0.267, which is smaller than 0. So the lottery is random. We used following calculation to estimate the p-value: 
$$\hat{p} =\frac{\sum T(X) \geqslant 0}{B}$$

```{r}
# calculate the p-value
# null hypothesis: T>=0, hv tendancy
p_val = mean(myboot$t>=0, na.rm = TRUE)
p_val
```

Here p-value is smaller than significant level $\alpha=0.05$, it also shows the data is random. So it seems the linear relation of lottery is not exist.

## 4. Tests the hypothesis

Implement a function depending on *data* and *B* that tests the hypothesis:

- *H0*: Lottery is random

- *H1*: Lottery is non–random

by using a permutation test with statistics `T`. The function is to return the p–value of this test. Test this function on our data with `B = 2000.

```{r}
set.seed(123456)
permutation_test <- function(B, data1){
  n = dim(data1)[1]
  stat = numeric(B)
  for(b in 1:B){
    # randoming X
    Gb = sample(data1$Day_of_year, n)
    data11 <- data1
    data11$Day_of_year <- Gb
    stat[b] <- stats(data11,1:n)
  }
  return(stat)
}

stat <- permutation_test(B=2000, data1)
stat0 <- stats(data1,1:nrow(data1))

re = data.frame( stat0 ,mean(abs(stat)>=abs(stat0)) )
colnames(re) <- c("average_statistic","p_value")
re

```
Since the the p-value is larger than 0.05, we cannot reject the null hypothesis. 

\newpage
## 5.Make a crude estimate

Make a crude estimate of the power of the test constructed in Step 4:

(a) Generate (an obviously non-random) dataset with n = 366 observations by using same X as in the original data set and $Y (x) = \max(0, \min(\alpha x + \beta, 366))$, where $\alpha = 0.1$
and $\beta\sim N(183, sd = 10)$.

(b) Plug these data into the permutation test with B = 200 and note whether it was
rejected.

(c) Repeat Steps 5a-5b for $\alpha = 0.2,0.3,...,10$.

What can you say about the quality of your test statistics considering the value of the
power?

```{r}
set.seed(123)
ce <- function(a, plot=FALSE){
  ## step (a)
  n <- 366
  b <- rnorm(n, mean=183, sd = 10)
  X <- data1$Day_of_year
  Y <- sapply(1:n,function(i){
    t <- min(a*X[i]+b[i],366)
    t <- max(0,t)
  })
  data15 <- data.frame(Day_of_year=X,Draft_No=Y)
  if(plot){
    plot(data15)
  }
  ## step (b)
  stat <- permutation_test(B=200, data15)
  
  stat0 <- stats(data15,1:n)
  re = data.frame( stat0 ,mean(abs(stat)>=abs(stat0)) )
  options(digits = 5)
  colnames(re) <- c("test_statistic","p_value")
  return(re)
}

ce(a=0.1, plot=TRUE)
ce(a=10, TRUE)
```
Since the p-values here are smaller than 0.05 and test-statistics are larger than 0, so we can reject the null hypothesis directly. Obviously, the data is non-random.

```{r echo=F}
## step (c)
b <- c()
a <- seq(0.2,10,0.1)
for (i in 1:length(a)) {
  re = ce(a[i])
  b[i] <- re$test_statistic
}
result <- data.frame(alpha=a, statistic=b)
ggplot(data = result,aes(x=alpha,y=statistic))+
  geom_point()
```

Based on the plot of alpha and test-statistic, we know that the test statistic we use is very strong. It expresses the non-randomness of our data and tends to 1.










\newpage
# Question 2: Bootstrap, jackknife and confidence intervals

```{r, echo=FALSE, results='hide'}
rm(list = ls())
data2 = read.csv2("prices1.csv")
```

The data you are going to continue analyzing is the database of home prices in Albuquerque, 1993. The variables present are `Price`; `SqFt`: the area of a house; `FEATS`: number of features such as dishwasher, refrigerator and so on; `Taxes: annual taxes paid for the house. Explore the file *prices1.xls*.

## 1. Plot the histogram of Price. 
Does it remind any conventional distribution? Compute the mean price.

```{r, echo=FALSE, warning=FALSE, out.height="300px"}
# calculate the mean of the price
price_mean = mean(data2$Price) # 1080.473

# create the plot - hist + distribution + mean  
ggplot(data = data2, aes(x = Price)) + 
  #geom_histogram(color = "#33CCFF", fill = "#33CCFF") + 
  ggtitle("Histogram of Price") + 
  geom_vline(aes(xintercept = price_mean, color = "Mean"),size=1.5) + 
  geom_histogram(aes(y=..density..),
                 colour="black",
                 fill="white",
                 bins=30)+
  geom_density(alpha=.2, fill="#FF6666")  
  # geom_density(kernel = "poisson") create a possion distribution
```

The present distribution looks like a possion or gamma distribution. 

The mean of the price is:
```{r}
price_mean
```

## 2. Bootstrap

Estimate the distribution of the mean price
of the house using bootstrap. 
Determine the bootstrap bias-correction and the variance of the mean price. Compute a 95% confidence interval for the mean price using bootstrap percentile, bootstrap BCa, and first-order normal approximation.

Estimate the distribution of the mean price of the house using bootstrap:
```{r}
# function - estiamte mean
stat1 <- function(vec,vn){
    return(mean(vec[vn]))
}
B=1000

set.seed(123456)
res1 = boot(data2$Price, stat1, R=B)
res1
```

According to the slide, we know that the bias-corrected estimator is
$$T_1:=2T(D)-\frac{1}{B}\sum^{B}_{i=1}T^{*}_i ~,$$
and variance of estimator by bootstrap is
$$\widehat{Var[T(.)]} = \frac{1}{B-1} \sum_{i=1}^B(T(D_i^*) - \overline{T(D^*)})^2.$$
Determine the bootstrap bias-correction and the variance of the mean price:
```{r}
# bias correcred estimator
2*res1$t0-mean(res1$t)

# variance of mean price (output of statistic)
var_boot <- 1/(B-1)*sum((res1$t-mean(res1$t))^2 )
var_boot
```

Compute a 95% confidence interval for the mean price using bootstrap percentile, bootstrap BCa, and first-order
normal approximation
```{r, echo=FALSE}

# default is a 95% confidence interval
ci <- boot.ci(res1, type = c("perc", "bca", "norm"))
print(ci) 

plot(res1)
```

## 3. Jackknife

Estimate the variance of the mean price using the jackknife and compare it with the
bootstrap estimate.

- Jackknife (n = B)

$$\widehat{Var[T(.)]} = \frac{1}{n(n-1)}\sum_{i=1}^n ((\rm T_i^*)- \rm J(T))^2,$$
where 
$$ T_i^* = nT(D)-(n-1)T(D_i^*)$$
and 
$$ J(T)= \frac{1}{n}\sum_{i=1}^{n}T_i^* $$


```{r}
# other jackknife function
# library(bootstrap)
# res2 <- jackknife(x=data2$Price,stat1)

n = nrow(data2)
constant = 1/(n*(n-1))

T_i_star = sapply(1:n, function(i){
  n * mean(data2$Price) - (n-1) * mean(data2[-i,1])
})

J_T = (1/n) * sum(T_i_star)

Var_T_jackknife = constant * sum((T_i_star - J_T)^2)
Var_T_jackknife
```

```{r echo=FALSE}
table = data.frame(boostrap = var_boot, jackknife= Var_T_jackknife)
knitr::kable(table)
```

It seems that bootstrap here is better than jackknife because of its smaller variance.


## 4. Compare the confidence intervals 

obtained with respect to their length and the location of the estimated mean in these intervals.

```{r echo=FALSE}
percent<-ci$percent
bca<-ci$bca
norm<-ci$normal
intervals = rbind(percent[4:5],bca[4:5],norm[2:3])

Len.CI=intervals[,2]-intervals[,1]
Esti.mean=intervals[,1]+Len.CI/2
table2 <- data.frame(Len.CI,Esti.mean)
rownames(table2) <- c("percent","bca","norm")
knitr::kable(table2)
```

On the one hand, percentile has the largest confidence intervals, but BCa and normal approximation have similar confidence intervals. On the other hand, the interval-means of percentile and BCa are similar ($\approx 1084$), but of normal approximation is around 1079.





\newpage

# Appendix
```{r ref.label=knitr::all_labels(), echo = T, eval = F}
```